Fail Faster

### 2025-03-14 paper/paper.tex
rewrite rng motiv section

% \subsubsection{Function call overhead}
\subsubsection{Choice of RNG}

\jwc{We really need to have explained the effect ordering equivalence before here.}

\jwc{
  \begin{itemize}
    \item Because the RNG sits at the heart of the generation hot path, RNG
    speed matters enormously. Which RNG you pick matters...
    \item Different PBT libraries use different RNGs --- QuickCheck and \bq use
    splittable random, Scalacheck uses \href{https://burtleburtle.net/bob/rand/smallprng.html}{this one}.  These
    are reasonably performant RNGs. Because this is the
    core of the hot path of the generator, speeding up sampling just a little can
    can dramatically speed up generation time.
    \item Significantly faster RNGs that still pass pRNG benchmarks like BigCrush exist (like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}), and using them can 
    speed up generation enormously --- in our microbenchmarks, the lehmer generator runs 4x faster than splittable random.
    \item However, showing this is tricky, because different RNGs have different sampling orders, but our
    eval is made much stronger by the fact that all of our generators are functionally equivalent.
    \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
    slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
    just how much you can improve bug-finding performance simply by speeding up the RNG.
  \end{itemize}
}

The core of any PBT generator library is a random-number generator. Following
the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented as an OCaml library called \texttt{Splittable\_random}. The precise
details of how SplitMix works are unimportant for the present paper, but the
main operations are defined in terms of bitwise operations on 64-bit integers.
However, due to implementation details related to the garbage collector, all
runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
is a pointer to a block of memory that contains a 64-bit integer.  This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output.

This has a significant impact on the performance of generators. A single call to
one of the \bq library functions may sample from \texttt{Splittable\_random}
multiple times, and each sample from \texttt{Splittable\_random} allocates 9
words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
will see in Section~\ref{section:eval} that this amount of allocation has a
significant impact on significant.

In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
to the same effect without leaving OCaml.

It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
the abstraction overhead of the generator DSL itself, but simply due to language
implementation and algorithm choices. We include this discussion --- and
the corresponding evaluation
--- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
should be engineered for performance.

\jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}

>>>>

% \subsubsection{Function call overhead}
\subsubsection{Choice of RNG}
\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}

The core of any PBT generator library is a (pseudo) random number generator.
Different PBT libraries use a wide range of different RNGs.
Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

% \jwc{There are a few aspects of PBT that make it interesting when it comes to PBT choice.
% \begin{itemize}
%   \item Doesn't need to be cryptographic.
%   \item doesn't even \emph{necessarily} need to pass sophisticated staistical
%   tests.  \cn
% \end{itemize}
% }

Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
As we will show in Section~\ref{eval}, speeding up sampling can dramatically speed up generation time and bugfinding speed.

We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
an impediment towards better bugfinding ability.

Significantly faster RNGs (that still pass pRNG benchmarks like BigCrush) exist like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}, and using them can 
speed up generation tremendously --- in microbenchmarks, a C implementation of the lehmer generator runs 4x faster than a C implementation of splitmix.

Similarly, PBT libraries could even eschew the requirement that a source of entropy pass statistical tests like bigcrush \cn.
This is common practice in testing already: fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy!
Bumping a pointer and reading from a buffer (which can be pipelined trivially) is likely to be significantly faster than any RNG with data-dependent arithmetic instructions.

In the context of this paper, however, demonstrating the benefits of any of the above options is tricky.
Our bug-finding evaluation's power is predicated on the fact that all of the different versions of a generator we test are extensionally identical:
for a given seed and size parameter, they all produce the \emph{same} value.
Using a different random number generator or different source of entropy would break this property.

In Section~\ref{eval}, we demonstrate that faster RNG matters for bugfinding by exploting a ``natural experiment'':
OCaml's \texttt{Splittable\_random} library is slow in a way that can be improved \emph{without} changing its extensional behavior,
namely that it uses \emph{boxed} 64-bit integers. By building a version that uses much faster unoxed 64-bit integer arithmetic, we can
demonstrate just how much bug-finding performance can be improved just by using a more performant RNG.

% \jwc{
%   \begin{itemize}
%     \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
%     slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
%     just how much you can improve bug-finding performance simply by speeding up the RNG.
%   \end{itemize}
% }

%  The precise
% details of how SplitMix works are unimportant for the present paper, but the
% main operations are defined in terms of bitwise operations on 64-bit integers.
% However, due to implementation details related to the garbage collector, all
% runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
% As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
% is a pointer to a block of memory that contains a 64-bit integer.  This means
% that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
% allocate memory cells to contain their output.

% This has a significant impact on the performance of generators. A single call to
% one of the \bq library functions may sample from \texttt{Splittable\_random}
% multiple times, and each sample from \texttt{Splittable\_random} allocates 9
% words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
% will see in Section~\ref{section:eval} that this amount of allocation has a
% significant impact on significant.

% In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
% performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
% We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
% to the same effect without leaving OCaml.

% It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
% the abstraction overhead of the generator DSL itself, but simply due to language
% implementation and algorithm choices. We include this discussion --- and
% the corresponding evaluation
% --- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
% should be engineered for performance.

% \jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}
